---
title: "Divvy DataSet"
output: html_notebook
---
Introduction: This was my first time reaching the Memory limit, with more than a gb of data and without being able to work it with Python without 
using chunks or doing some "tricks", however R came to save me and was able to work it easily. Both are strong tools but I prefer python, maybe I
should use Polar for working and leave Pandas as a lot of Data Scientists decided to.

If you need to use this work, feel happy to copy or use it to learn by yourself!
Greets, 
Jlob

#Install Packages for working

```{r}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
library(tidyverse)
library(lubridate)
library(ggplot2)
```
### Data Collection 

```{r}

q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")

```

### Checking Structure of data
```{r}
print('Q2 2019 Data')
colnames(q2_2019)
```

```{r}
print('Q3 2019 Data')
colnames(q3_2019)
```

```{r}
print('Q4 2019 Data')
colnames(q4_2019)
```

```{r}
print('Q1 2020 Data')
colnames(q1_2020)
```

### Fixing Columns inconsistencies, taking reference lastest edition as new format.

```{r}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))

```

### Checking if dtypes are correct to join them together.

```{r}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```

### Converting numbers to characters, to merge together.

```{r}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

```

### Stacking together

```{r}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

### Removing redundancies from older formats (Using -c to keep all others columns.)

```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))

```

### Inspecting health of the new dataset

```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics

```
### Fixing names of members types to work with new format, the error has been seen while inspecting the data with view()

```{r}
unique(all_trips$member_casual)
table(all_trips$member_casual)

```
### Fixing names
```{r}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))

```

### Checking if fixed

```{r}
table(all_trips$member_casual)
```
### Adding columns that will help us in the analysis of data
### Helpful resource to do this (https://www.statmethods.net/input/dates.html)
```{r}

all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

### Adding calculated column (Ride_Length)
### Resource to do this(https://stat.ethz.ch/R-manual/R-devel/library/base/html/difftime.html)

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

### Checking for the result

```{r}
str(all_trips)
```
### Ride_length has a format that I do not need.

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)

```
### Final checking before cleaning

```{r}
str(all_trips)
```
### Cleaning from bad data with emptys and negative values, seen by inspecting the data
### Good practice is to do not replace the database and create a new version.
### Useful Resources (https://www.datasciencemadesimple.com/delete-or-drop-rows-in-r-with-conditions-2/)

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
summary(all_trips_v2)

```

### Descriptive analysis, but it can be done with summary too.

```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
summary(all_trips_v2$ride_length)
```

### Compare stats for type of members

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```

###Comparing statistics per day and per type of customer.

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

### Not sorted

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

###

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

# analyze by type of member and day

```{r}
df <- all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
df
```

#Visualize

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  #EXtra step to visualize
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```

# Visualization for average duration and weekday

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```
### Exporting
### Resources :  https://datatofish.com/export-dataframe-to-csv-in-r/
```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = 'report.csv')
write.csv(df, file= "report2.csv")

```



